DOpElib: A Goal Oriented Software Library for Computing PDEs and Optimization Problems

### 2012-11-05 Paper_May_2012/main.tex
Added comments on Stucture for Instat and OPT added Example on compliance minimization.

\subsubsection{Nonstationary problems}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi) = \int_I f(phi) \quad \forall\phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\[
\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt
+ (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\]
for all $\phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J(\cdot)$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
a global object that treats the routines
in DOpElib:
\begin{lstlisting}
  template<typename Triangulation, typename FE, typename times>
      void MethodOfLines_SpaceTimeHandler(Triangulation &, FE, times);
\end{lstlisting}
Spatial discretization is the same
as discussed previously, and thus, we briefly demonstrate
routines to perform temporal integration in the following.

\subsubsection{Problem-dependent 
implementation of nonstationary problems}
\label{sec:timedep:implementation}
We explain in more detail the temporal discretization
and specific issues of its implementation. 
To keep the presentation easy, we abuse notation 
by neglecting
all material coefficients
and give formal-related idea based on the weak 
form of the Navier-Stokes equations:
Find $(v,p)$ such that
\begin{align*}
(\partial_t v,\phi) 
+ (\nabla v, \nabla \phi)
+ (v\cdot\nabla v,\phi)
-(p,\nabla\cdot \phi)
+(\nabla\cdot v, \chi)
=(f,\phi).
\end{align*}
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $(v^n,p^n)$, 
find $(v^{n+1}, p^{n+1})$ (and multiplication 
through by the timestep $k$):
\[
(v^{n+1} - v^{n}, \phi)
+ k\theta (\nabla v^{n+1}, \nabla \phi)
+ k\theta (v^{n+1}\cdot\nabla v^n + 
  v^{n}\cdot\nabla v^{n+1},\phi)
- k (p^{n+1},\nabla\cdot \phi)
+ (\nabla\cdot v^{n+1}, \chi)
= k\theta (f^{n+1},\phi) + k(1-\theta) (f^{n},\phi)
- k(1-\theta) (\nabla v^{n}, \nabla \phi) 
\]
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}

@@

\subsection{Solving of optimization problems}\label{sec:opt}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
As before, the main part is done in the \texttt{main} function there is no
great difference to the previous explications. However, 
since we utilize gradient-based optimization routines in the philosophy of
\cite{Troe99,HiKu01},
all derivatives must be implemented. All this information
must given to the \texttt{localpde.h} file. Herein, one has to 
implement the adjoint, tangent and adjoint Hession equations. 
The algorithmic parts are described in the literature \cite{BeMeVe06}.

Concretely, let us focus on the Laplace equation and 
the following optimization problem (taken from 
Example 2) in \texttt{Examples/OPT/StatPDE}:
\begin{align*}
\min_{(q,u)\in \mathbb R^3 \times H_0^1(\Omega; \mathbb R^2)} J(q,u) &=
\frac{1}{2} \sum_{i=0}^2 |(u-\overline u)(x_i)|^2 + \frac{\alpha}{2}\|q\|^2\\
\text{s.t.} (\nabla u,\nabla \phi) &= (f(q),\phi)\;\;\forall\,\phi \in H^1_0(\Omega; \mathbb R^2)
\end{align*}
on the domain $\Omega = [0,1]^2$, with
\begin{itemize}
\item the observation points
\begin{align*}
x_0 = (0.5, 0.5), \quad x_1 = (0.5, 0.25),\quad x_2 = (0.25, 0.25),
\end{align*}
\item the regularization parameter $\alpha = 0$, 
\item the right hand side
\begin{align*}
 f(q) &= q_0 \left(\begin{matrix}2\pi^2  \sin( \pi x) \sin(\pi y)\\0 \end{matrix}\right)\\
      &+ q_1 \left(\begin{matrix}5\pi^2  \sin( \pi x) \sin(2\pi y)\\0 \end{matrix}\right)\\
      &+ q_2 \left(\begin{matrix}0 \\8\pi^2  \sin(2\pi x) \sin(2\pi y)\end{matrix}\right)\\
\end{align*}
\item and the exact solution given by 
\begin{align*}
 \overline{q} &= \bigl(1;0.5;1\bigr)\\
 \overline{u}& = \left(\begin{matrix} \sin( \pi x)( \sin(\pi y)+0.5\sin(2\pi y))\\\sin(2\pi x) \sin(2\pi y) \end{matrix}\right).
\end{align*}
\end{itemize}
\begin{defi}[Constrained optimization]
\label{problem_constraint_optimization}
Minimize the cost functional $J(q,u)$ subject to the state equation
$a(q,u)(\phi) = 0$ for $(q,u) \in 
Q\times  \{v_D+{\cal X}\}$. 
\end{defi}
The constrained optimization problem  on the space $Q_d\times {\cal X}$ 
is reformulated into an unconstrained optimization problem on the
space $Q_d$. Therefore, we assume the existence of the solution operator  
$S:Q_d\rightarrow {v_D+{\cal X}}$ with a unique solution
$U=S(q)$. Herewith, we define the reduced cost functional
$j:Q_d\rightarrow \mathbb{R}$ by 
\begin{equation}
  \label{opt_reduced_functional}
  j(q) := J(q,S(q)) .
\end{equation}
Thus, the constrained optimization problem can be formulated by means
of  
\begin{defi}[Unconstrained optimization]
  \label{opt_reduced_problem}
  Minimize $j(q)$ for $q\in Q_d$.
\end{defi}
Newton's method to solve Problem~\ref{opt_reduced_problem} reads:
For $l=0,1,\ldots,$ solve
\begin{equation}
\label{opt_newton_method}
\begin{aligned}
j''(q^l)(\delta q , \tau q) &= -j'(q^l)(\delta q) 
\quad\forall\tau q\in Q_d, \\
q^{l+1} &= q^l + \omega\delta q,
\end{aligned} 
\end{equation}
with a line search parameter $\omega\in (0,1]$ which 
will are implemented in \texttt{reducednewtonalgorithm.h}.
Specifically, the residual and the Hessian of Newton's method 
\eqref{opt_newton_method}
can be computed with the help of the following two results:
\begin{proposition}[Residual of Newton's method]
\label{prop_newton_residual}
Let $q\in Q_d, u=S(q)\in {\cal X}$
and $z\in {\cal X}$ the dual solution be obtained after solving 
the first and second equation of the KKT system 
\eqref{opt_KKT_system}. 
Then the residual of Newton's method \eqref{opt_newton_method} 
is defined as
\begin{equation*}
j'(q)(\delta q) := {\cal L}'_{q}(q,u,z)(\delta q),
\end{equation*}
i.e., in explicit representation
\begin{equation*}
j'(q)(\delta q) := \alpha_T (q , \tau q)_Q - 
A_{q}'(q,u)(\tau q , z).
\end{equation*}
\end{proposition}
\begin{proof}
The proof uses standard techniques. Details can be found in  
\cite{BeMeVe06}.
\end{proof}
\begin{proposition}[Hessian of Newton's method]
  \label{prop_newton_hessian}
  Let $q\in Q_d, u=S(q)\in {\cal X}$ and $z\in {\cal X}$ and  $\delta
  q\in Q_d$ be given. Further, let $\delta u\in {\cal X}$ be the
  solution of the tangent problem 
  \[
  \delta u\in{\cal X},\quad 
  a_{u}'(q,u)(\delta u , \phi) 
  = -a_{q}'(q,u)(\delta q , \phi)  \quad\forall \phi\in {\cal
    X},
  \]
  and $\delta z\in {\cal X}$ the solution of the adjoint Hessian 
  problem  
  \[
  a_{u}'(q,u)(\phi, \delta z)
  = J''_{uu}(q,u)(\delta u, \phi)
  - a_{uu}''(q,u)(\delta u, \phi, z)
  - a_{qu}''(q,u)(\delta q , \phi, z)  \quad\forall \phi\in
  {\cal X}. 
  \]
 Then it holds  for all $\tau q\in Q_d$:
  \[
  j''(q)(\delta q, \tau q):= 
  \alpha_T (\delta q , \tau q)_Q -
  a_{qq}''(q, u)(\delta q, \tau q , z) -
  a_{uq}''(q,u)(\delta u, \tau q , z) -
  a_{q}'(q,u)(\tau q , \delta z).
  \]
\end{proposition}
\begin{proof}
  For the proof this statement (including its time-dependent version), we refer to 
  Becker et al. \cite{BeMeVe06}.
  The stationary version of this proposition is easily derived by neglecting 
  all time derivatives, which concludes the assertion. 
\end{proof}
The implementation of all previous terms 
is realized in the \texttt{localpde.h} file.
As an example, we examine the adjoint 
\begin{equation}
  \label{opt_KKT_system}
  \begin{aligned}
    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X},
  \end{aligned}
\end{equation}

The implementation it split into equation terms, i.e., 
\begin{lstlisting} 
  void CellEquation_U(...)
   {
     // implement (\nabla \phi, \nabla z)
   }
\end{lstlisting}
and functional terms, i.e.,     
\begin{lstlisting} 
  virtual void PointValue_U(...)
      {
        // implement here the derivative of 
        // the point cost functional
      }
\end{lstlisting}
All remaining terms of the previous two propositions 
are implemented in the same manner.

Afterwards, it remains to use an appropriate 
optimization algorithm in order to solve the 
problem. At present we offer a selection of algorithms that solve the reduced optimization 
problem where the PDE constraint has been eliminated as explained previously.
\begin{itemize}
\item \texttt{opt\underline{ }algorithms/reducedalgorithm.h} An interface for all 
  optimization problems in the reduced formulation. It offers some test functionality
  to assert that the derivatives of the problem are computed correctly.
\item \texttt{opt\underline{ }algorithms/reducednewtonalgorithm.h}
  A line-search Newton algorithm using a cg method to invert the reduced hessian. 
  Implementation ignores any additional constraints.
\item \texttt{opt\underline{ }algorithms/reducedtrustregionnewton.h}
  A trust region Newton algorithm using a cg method to invert the reduced hessian.
  Implementation ignores any additional constraints.
\item \texttt{opt\underline{ }algorithms/reduced\underline{ }snopt\underline{ }algorithm.h}
  An algorithm to solve reduced optimization problems with additional control constraints.
  ((reduced) state constraints are not yet implemented.)
\item \texttt{opt\underline{ }algorithms/reducednewtonalgorithmwithinverse.h}
  Line-search Newton algorithm that assumes there exists a method in the reduced problem
  that can invert the reduced hessian. (This usually makes sense only if there is no 
  PDE constraint.)
\item \texttt{opt\underline{ }algorithms/generalized\underline{ }mma\underline{ }algorithm.h}
  An implementation of the MMA-Algorithm for structural optimization using an augmented
  Lagrangian formulation for the subproblems. The subproblem is implemented using the 
  special purpose file
  \texttt{include/augmentedlagrangianproblem.h}.
\end{itemize}

\subsection{Independent features}
In this final section, we shall give a brief summary of features 
that are applicable (in general) to all types of problems. Specifically,
\dope{} provides goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method \cite{BeRa96}. Here, a certain 
cost functional such as a point evaluation, line intergration or domain
integration is considered. Determining the lowest error between the 
(unknown) exact solution and its numerical approximation for lowest computational
cost is the main goal of the DWR method.

Moreover, \dope{} provides standard error estimators based on residual evaluation and 
a ZZ error estimator, i.e., measureing the smoothness of the discrete first derivatives.
The latter one is accessed via \deal{}.

In addition, \dope{} provides multi-mesh procedures 
\todo{kann Winni hier etwas schreiben?}

Concerning temporal integration schemes, the library comes 
with various time stepping schemes (based on finite differences), 
such as the first-order forward Euler and first-order backward Euler scheme. 
To compute fully nonstationary processes, the second-order schemes
Crank-Nicolson (trapozoidal rule), 
shifted Crank-Nicolson (slight shift to the implicit side to maintain
global stability), 
and the Fractional-Step-$\Theta$ scheme, are implemented.

%\subsection{Problem description}
%In order to allow our algorithms the automatic assembly of all required 
%data we need to have some container which contains the complete problem 
%description in a common data format. For this we have the following 
%classes in \texttt{DOpEsrc/container}
%\begin{itemize}
%  \item \texttt{pdeproblemcontainer.h} Is used to describe  stationary PDE problems.
%  \item \texttt{instatpdeproblemcontainer.h} {\bf TODO is still missing but will be used for nonstationary problems.} This will be implemented once we have nonstationary optimization problems running to avoid error duplication in the coding process.
%  \item \texttt{optproblem.h} Is used to describe  OPT problems governed by 
%    stationary PDEs. 
%  \item \texttt{instatoptproblemcontainer.h} Is used to describe  OPT problems
%    governed by nonstationary PDEs. The only difference to the stationary case
%    is that we need to specify a time-stepping method.  
%  \item \texttt{interfaces/functionalinterface.h} This gives an interface 
%    for the functional $J(\cdot)$ and any other functional you may want to evaluate.
%    In general this can be used as a base class to write your own functionals 
%    in examples. We note that we only need to write the integrands on 
%    elements or faces the loop over elements will be taken care of in the integrator.
%    Specifically, derivatives are written therein, too.
%\end{itemize}
%In order to fill these containers there are two things to be done,
%first we need to actually write some data, for instance,
%the semilinear form $a(\cdot)(\cdot)$, a target functional $J(\cdot)$, etc.,
%which describe the problem. Then we have to select some numerical 
%algorithm components like finite elements, linear solvers $\ldots$.
%The latter ones should be written such that when exchanging these components
%none of the problem descriptions should require changes. 
%Note that it still may be necessary to write some additional descriptions, 
%e.g., if you solve the PDE with a fix point iteration you don't need derivatives
%but if you want to use Newton's method, derivatives are needed.
%We will start by discussing the problem description components implemented so far
%\subsection{Numeric components}
%These are the components from which a user needs to select some in order to actually 
%solve the given problem. They will not require any rewriting, but sometimes it is 
%advisable to write other than the default parameter into the param file for the 
%solution.
%\subsubsection{Space-time handler}
%First we need to select a method how to handle all dofs in space and time.
%\begin{itemize}
%\item \texttt{basic/spacetimehandler\underline{ }base.h} This class is used to define 
%  an interface to the dimension independent functionality of all space time dof handlers.
%  {\bf TODO: Beispiele geben}
%\item \texttt{basic/statespacetimehandler.h} Another intermediate interface class which adds 
%  the dimension dependent functionality if only the variable $u$ is considered, i.e., a 
%  PDE problem.
%\item \texttt{basic/spacetimehandler.h } Same as above but with both $q$ and $u$, i.e., for
%  OPT problems.
%\item \texttt{basic/mol\underline{ }statespacetimehandler.h} Implementation of a method of 
%  line space time dof handler for PDE problems. It has only one spatial 
%  dofhandler that is used for all time intervals.
%\item \texttt{basic/mol\underline{ }spacetimehandler.h} Same as above for OPT problems.
%  A separate spatial dof handler for each of the variables $q$ and $u$ is maintained 
%  but only one triangulation.
%\item \texttt{basic/mol\underline{ }multimesh\underline{ }spacetimehandler.h}
%  Same as above, but now in addition the triangulations for $q$ and $u$ can be refined
%  separately from one common initial coarse triangulation. Note that this will
%  in addition require the use of the multimesh version for integrator and 
%  face- as well as celldatacontainer.
%\end{itemize}
%Note that we use these for stationary problems as well, but then you don't have to specify
%any time information.
%\subsubsection{Container classes}
%Second you will need to specify some container classes to be used to 
%pass data between objects. At present you don't have much choice, but you may wish 
%to reimplement some of these if you need data that is not currently included in 
%the containers.
%\begin{itemize}
%\item \texttt{container/celldatacontainer.h} This object is used to pass data 
%  given on the current element (cell) of the mesh to the functions in PDE, functional, 
%  $\ldots$. 
%\item \texttt{container/facedatacontainer.h} This object is used to pass data 
%  given on the current face of the mesh to the functions in PDE, functional, 
%  $\ldots$. 
%\item \texttt{container/multimesh\underline{ }celldatacontainer.h} This is the same as the 
%  celldatacontainer, but it
%  is capable to handle data defined on an alternative triangulation.
%\item \texttt{container/multimesh\underline{ }facedatacontainer.h} This is the same as the
%  facedatacontainer, but it
%  is capable to handle data defined on an alternative triangulation.
%\item \texttt{container/integratordatacontainer.h} This contains some data that 
%  should be passed to the integrator like quadrature formulas and the above cell and 
%  face data container.
%\end{itemize}
%\subsection{Reduced problems (Solve the PDE)}
%At times it is nice to remove the PDE constraint in (OPT). 
%This is handled by so called reduced 
%problems (for algorithmic aspects we refer the reader to 
%\cite{BeMeVe06}). 
%This means that the reduced problem implicitly solves the PDE whenever required
%and eliminates the variable $u$ from the problem.
%\begin{itemize}
%\item \texttt{reducedproblems/statpdeproblem.h} This is used to remove the variable $u$ in 
%  a stationary PDE problem. This means that call the method \\
%  \texttt{StatPDEProblem::ComputeReducedFunctionals} will evaluate the functionals 
%  defined in the problem description, i.e., in \texttt{PDEProblemContainer}, in the 
%  solution of the given PDE.
%\item \texttt{reducedproblems/statreducedproblem.h} This eliminates $u$ from the OPT
%  problem with a stationary PDE.
%\item \texttt{reducedproblems/instatreducedproblem.h} The same as above but for a
%  nonstationary PDE. {\bf FIXME there is something wrong in this file see FIXME 
%    comment in the source.}
%\item \texttt{reducedproblems/voidreducedproblem.h} A wrapper file that eliminates $u$ 
%  if it is not present anyways. This is used so that we can use the same routines to 
%  solve problems that have no PDE constraint.
%\end{itemize}

%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\subsection{Interfaces to other software packages}
%Since DOpE uses as basis the software library \deal{}, all its interfaces
%to other packages can be accessed to. For instance, such libararies 
%are trilinos for with an algebraic multigrid method, MPI for a parallel
%solution of the problem.
%
%In addition DOpE itself has an interface to SNOPT.
%
%%\section{Documentation, Code Development, Webpage}
%%\label{documentation}
%%At the present step, the \dope{} project comes with 
%%an detailed documentation of all features and examples 
%%in pdf format and an detailed programming code 
%%documentation. 
%%
%%\todo{License? How to access to DOpE?}
%%\todo{Name of the website}
%%
%%In addition, the \dope{} test suite provides some regression tests. 
%%They are run to compare 
%%the output to previous outputs. This is useful (necessary) after 
%%changing programming code anywhere in the library. If a test
%%succeeds, everything is fine in the library. If not, you should not
%%check in your code into DOpE. Please make sure what is going wrong and WHY!
%%Every command is computed via a Makefile. 
%%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{applications}
We present three numerical tests which demonstrate different
features of \dope{}:
\begin{itemize}
\item In the first numerical example, we consider 
goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method for the stationary Navier-Stokes equations.
\item The second example presents nonstationary fluid-structure 
interaction. The challenges for those kind of problems are the multi-domain
character of the coupled problem and the treatment of the coupling conditions.
The problem is formulated in a variational monolithically-coupled way which 
allows to consider goal-oriented mesh refinement and gradient-based optimization.
\item In the third numerical test, an optmization problem for structural mechanics
is discussed.
\end{itemize}

@@

Details 
on parameters and evaluation functionals and other results 
can be found in \cite{HrTu06b,BuSc06,DeHaeAnnBrVie10,Wi11}. 
The Fractional-Step-$\theta$ scheme is used for time discretization with
different time step sizes $k$. However, the time stepping scheme can be 
very easily chosen in the \texttt{main.cc} function by declaring
\begin{lstlisting}
#define TSP1 ForwardEulerProblem
#define TSP2 BackwardEulerProblem
#define TSP3 CrankNicolsonProblem
#define TSP4 ShiftedCrankNicolsonProblem
#define TSP5 FractionalStepThetaProblem
\end{lstlisting}
and then using in the code the designated scheme, 
for example here the Fractional-Step-$\theta$ scheme
\texttt{TSP5}.

@@

\subsection{Compliance minimization}
Winni
Results already obtained.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{conclusions}
In this article, we described the features 
and applications of the \dope{} project, which 
was initiated at the Heidelberg University in 2010.
Specifically, the applications cover a broad 
spectrum of numerical exmamples motivated 
by different disciplines. 
Currently, we are developing concepts for the efficient 
numerical solution of time-dependent optimization
problems governed by PDEs.

\section{Outlook}
\todo{Was wollen wir noch in den nächsten 1-2 Jahren erreichen?}

>>>>

\subsubsection{Nonstationary problems}\label{sec:timedep}
To describe the solution of nonstationary 
PDEs, we consider a time-dependent semi-linear form
Let $I:=[0,T]$ be a time interval with end time point value $T$.
To find some $u$ in a weak form, we define
\[
\int_I a(u)(\phi) = \int_I f(\phi) \quad \forall\phi\in X,
\]
with some suitable Bochner space $X$.  As model problem,
we consider the nonstationary Navier-Stokes equations
in a domain $\Omega\in\mathbb{R}^d$. Find: 
$(v,p)\in X$ with
\[
\int_I \bigl[ (\partial_t v, \phi)
+ \nu (\nabla v, \nabla \phi) + (v\cdot\nabla v, \phi)
- (p,\nabla\cdot \phi)
+ (\nabla\cdot v, \chi)\bigr] \, dt
+ (v(0) - v^0, \phi(0))
= \int_I (f,\phi) \, dt 
\]
for all $\phi:= (\phi, \chi) \in X$ with
\[
X:= \bigl\{ 
(v,p) | v\in L^2(I,H_0^1(\Omega)^d) , 
\partial_t v\in L^2(I, L^2(\Omega)^d), 
p\in L^2(I,L^2(\Omega)\setminus\mathbb{R} \bigr\}.
\]
As for stationary problems, the goal is to compute
some target functional $J(\cdot)$ which is now possibly 
time-dependent. To organize the discretization 
process, we need to perform spatial and temporal 
discretization. This is again done via
an object that takes care of organizing the degrees of 
freedom, as in the case of stationary problems.
The only difference to the stationary case is that we need to 
add the temporal subdivision \texttt{times}. In the simplest 
case, when we have the same spatial triangulation 
at all times, the constructor could look as follows:
\begin{lstlisting}
  template<typename Triangulation, typename FE, typename times>
      void SpaceTimeHandler(Triangulation &, FE, times&);
\end{lstlisting}

Now we may discretize the problem with some time-stepping scheme, which will 
remove the temporal derivative and replace it by some difference quotients. 
Since the exact way of doing this depends on the time-stepping scheme 
we need a new object that takes the user-given description of the PDE and 
generates the appropriate combinations of spatial integrals and scaling.
Thus a generic problem may look as follows:
\begin{lstlisting}
  template<typename PROBLEM>
      class TimeStepProblem;
\end{lstlisting}
At present we have prepared the following time stepping schemes in \dope:
\begin{itemize}
\item Forward-(Explicit)-Euler
\item Backward-(Implicit)-Euler
\item Crank-Nicolson scheme
\item shifted Crank-Nicolson scheme (to deal with singular initial data)
\item fractional-step-$\theta$ scheme
\end{itemize}
With this we can now run a loop over all timesteps, where in each of the 
timesteps we can reuse the functionality to solve stationary PDEs 
applied to the time-step problem.

Now it is apparent, why we introduced the \texttt{StatPDEProblem}, because
we now define the \texttt{InstatPDEProblem} to contain the necessary 
loop over all time points. Next, we will describe the methods the user needs to 
provide in addition to the stationary case.

\subsubsection{Problem-dependent 
implementation of nonstationary problems}
\label{sec:timedep:implementation}
We explain in more detail the temporal discretization
and specific issues of its implementation. 
To keep the presentation easy, we abuse notation 
by neglecting
all material coefficients
and give formal-related idea based on the weak 
form of the Navier-Stokes equations:
Find $(v,p)$ such that
\begin{align*}
(\partial_t v,\phi) 
+ (\nabla v, \nabla \phi)
+ (v\cdot\nabla v,\phi)
-(p,\nabla\cdot \phi)
+(\nabla\cdot v, \chi)
=(f,\phi).
\end{align*}
Temporal discretization using the One-step-$\theta$ scheme reads:
Given the old timestep solution $(v^n,p^n)$, 
find $(v^{n+1}, p^{n+1})$ (and multiplication 
by the inverse of the timestep $k$):
\begin{align*}
(v^{n+1} - v^{n}, \phi)
&+ k\theta (\nabla v^{n+1}, \nabla \phi)
+ k\theta (v^{n+1}\cdot\nabla v^n + 
  v^{n}\cdot\nabla v^{n+1},\phi)\\
&- k (p^{n+1},\nabla\cdot \phi)
+ (\nabla\cdot v^{n+1}, \chi)\\
=&\; k\theta (f^{n+1},\phi) + k(1-\theta) (f^{n},\phi)
- k(1-\theta) (\nabla v^{n}, \nabla \phi) 
\end{align*}
The concrete choice of $\theta$ leads to a
specific time stepping scheme. For instance,
$\theta = 0.5$ gives the Crank-Nicolson scheme,
\begin{lstlisting}
   CrankNicolsonProblem<...>
\end{lstlisting}

@@

\subsection{Solving of optimization problems}\label{sec:opt}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
To keep the presentation brief we will neglect the additional bounds
\[
a \le q \le b,\qquad g(q,u) \le 0
\]
although they can be treated by \dope as well.
Then, in order to minimize the functional $J$ one may apply the, so-called, 
reduced approach, in which we eliminate the dependent variable $u$ by the solution 
operator $S$ of the PDE, i.e., $a(q,S(q))(\phi) = 0 \quad \forall \phi\in V$. 
This leads to the problem
\begin{align*}
\min\;&j(q) = J(q,S(q)). 
\end{align*}
Then, one can apply any algorithm to solve the, unconstraint, minimization 
problem. This, will in any case require the evaluation of $j(q)$ which 
we can do, e.g., using \texttt{StatPDEProblem::ComputeReducedFunctional} for 
a stationary PDE. If one would like to use derivative based methods, one 
needs in adition derivatives of $j$ and thus of $J$ as well as of $S$. 
To do so, we remark, that one can calculate the derivative of 
$j$ using the adjoint calculus. Then the directional derivative 
of $j$ in direction $\delta q$ is given by
\[
j'(q)(\delta q) = J'_q(q,S(q))(\delta q) + (S^* J_u'(q,S(q)),\delta q)
\]
where the index $q$ or $u$ denote directional derivatieves for the 
respective argument. To calculate action of the adjoint operator $S^*$ 
we note that $z =  S^* J_u'(q,S(q))$
can be computed via the adjoint PDE
\begin{equation}
  \label{opt_KKT_system}
  \begin{aligned}
    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X}.
  \end{aligned}
\end{equation}
Similarly formulas for the second derivatives can be obtained.

To do so, we define an object 
\begin{lstlisting} 
  template<typename PDE, typename COSTFUNCTIONAL>
    class OPTProblemContainer;
\end{lstlisting}
that collects the user given PDE and functional descriptions and constructs the 
required PDE and right hand side information needed for the calculations.

This means for the user only the derivatives of the functional and the PDE 
needs to be provided in addition to the 
\begin{lstlisting} 
  void CellEquation_U(...)
   {
     // implement directional derivative of the CellEquation
   }
\end{lstlisting}
and functional terms, i.e.,     
\begin{lstlisting} 
  virtual void BoundaryValue_U(...)
      {
        // implement here the derivative of 
        // the boundary cost functional
      }
\end{lstlisting}

Due to the design of the classes it is easily possible to extend any given 
and running code for a PDE both to the nonstationary counterpart, if the PDE is 
stationary, as well as to an optimization problem where the constraint is 
given by this PDE. Meaning that the use of our framework for the simulation of 
PDEs will, due to the enforced interface, enable easy application of state-of-the-art 
optimization algorithms to minimization problems governed by the PDE.

\subsection{Additional features}
In this final section, we shall give a brief summary of features 
that are applicable (in general) to all types of problems. Specifically,
\dope{} provides goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method \cite{BeRa96}. Here, a certain 
cost functional such as a point evaluation, line intergration or domain
integration is considered. Determining the lowest error between the 
(unknown) exact solution and its numerical approximation for lowest computational
cost is the main goal of the DWR method. The same routines 
can also be used, without any modification, to use standard residual error estimators by exchanging the 
error estimator class.

Since in many applications it is desirable to use different discretizations
for the state and the control \dope{} provides multi-mesh support, meaning that
onec a problem has been written correctly using the same mesh for all finite element
variables one can switch to a different SpaceTimeHandler that allows different meshes for all 
variables without the need to change the user-given implementation of the PDE and functional.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Applications}
\label{applications}
We present three numerical tests which demonstrate different
features of \dope{}:
\begin{itemize}
\item In the first numerical example, we consider 
goal-oriented mesh refinement with the help of the 
Dual-Weighted-Residual (DWR) method for the stationary Navier-Stokes equations.
\item The second example presents nonstationary fluid-structure 
interaction. The challenges for those kind of problems are the multi-domain
character of the coupled problem and the treatment of the coupling conditions.
The problem is formulated in a variational monolithically-coupled way which 
allows to consider goal-oriented mesh refinement and gradient-based optimization.
\item In the third numerical test, an optmization problem for structural mechanics
is discussed.
\end{itemize}

@@

Details 
on parameters and evaluation functionals and other results 
can be found in \cite{HrTu06b,BuSc06,DeHaeAnnBrVie10,Wi11}. 
The Fractional-Step-$\theta$ scheme is used for time discretization with
different time step sizes $k$. However, the time stepping scheme can be 
very easily chosen in the \texttt{main.cc} function by choosing an appropriate 
time stepping scheme, at present the following are availiable.
\begin{lstlisting}
ForwardEulerProblem
BackwardEulerProblem
CrankNicolsonProblem
ShiftedCrankNicolsonProblem
FractionalStepThetaProblem
\end{lstlisting}

@@

\subsection{Compliance minimization}
In this example, we consider the compliance minimization of a standard MBB-Beam. 
The thickness variable is allowed to take intermediate value apart from $0$ and $1$,
i.e., we consider a variable thickness sheet. The discretization is done 
using $Q_2$ elements for the displacement and discontinuous $P_0$ elements for the 
thickness.
\begin{figure}
\centering
\includegraphics[width=1.\textwidth, viewport=150 0 1024 500, clip]{Pictures/MBB.png}
\caption{Thickness distribution for the MBB-beam minimum compliance problem.} 
\label{res:mbb}
\end{figure}
A typical thickness distribution is shown in Figure~\ref{res:mbb}.

\section*{Thanks}
\todo{Wem wir so danken wollen, Boris, Dominik, Wolfgang ...?}
